---
title: "Pierce County Property Valuation"
author: "B1_Pierce-House-Price"
format:
  revealjs:
    theme: serif  
    transition: slide
    slide-number: true
    toc: false
    toc-depth: 2
    incremental: false
execute:
  echo: false
  warning: false
  message: false
---
# Use Case & Problem
<style>
.reveal h1 {
  font-size: 1.5em !important;
}
.reveal h2 {
  font-size: 1.3em !important;
}
.reveal h3 {
  font-size: 1.0em !important;
}
</style>

- Predict **residential property prices** in Pierce County, WA  
- Use **public assessor data**: structure, location, utilities, quality  
- Support **fair transactions**, **policy**, and **risk assessment**  
- Build interpretable models for **price prediction**

# Project Objectives

- 🎯 **Target**: Achieve **<10% MAPE** for predictions  
- 🔍 **Key Drivers**: What most influences sale price?  
- 🗺️ **Regional Differences**: How does location impact value?  
- 💎 **Luxury Markers**: What defines high-end properties?  

```{r}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(tigris)
library(sf)
library(dplyr)
library(ggplot2)
library(viridis)
library(scales)
library(grid)
library(GGally)
library(tidyr)
library(corrplot)

full_data <- read_csv("data/final_data_incl_land_area.csv", show_col_types = FALSE)
```

```{r}
full_data <- full_data %>%
  dplyr::select(
    Sale_Date_Raw,
    Sale_Price_Raw,
    Square_Feet_Raw,
    Latitude_Raw,
    Longitude_Raw,
    Bedrooms_Raw,
    Bathrooms_Raw,
    Stories_Raw,
    Quality,
    Condition,
    Neighborhood,
    Street_Type,
    Utility_Water,
    Utility_Electric,
    Utility_Sewer,
    Improved_Vacant_Raw,
    Year_Built_Raw,
    Net_Land_Square_Feet_Raw = Net_Land_Square_Feet_Raw
  ) %>%
  mutate(
    Sale_Date_Raw       = as.Date(Sale_Date_Raw),
    Sale_Price_Raw      = as.numeric(Sale_Price_Raw),
    Square_Feet_Raw     = as.numeric(Square_Feet_Raw),
    Latitude_Raw        = as.numeric(Latitude_Raw),
    Longitude_Raw       = as.numeric(Longitude_Raw),
    Bedrooms_Raw        = as.numeric(Bedrooms_Raw),
    Bathrooms_Raw       = as.numeric(Bathrooms_Raw),
    Stories_Raw         = as.numeric(Stories_Raw),
    Quality             = as.factor(Quality),
    Condition           = as.factor(Condition),
    Neighborhood        = as.factor(Neighborhood),
    Street_Type         = as.factor(Street_Type),
    Utility_Water       = as.factor(Utility_Water),
    Utility_Electric    = as.factor(Utility_Electric),
    Utility_Sewer       = as.factor(Utility_Sewer),
    Improved_Vacant_Raw = as.factor(Improved_Vacant_Raw),
    Year_Built_Raw      = as.numeric(Year_Built_Raw),
    Net_Land_Square_Feet_Raw = as.numeric(Net_Land_Square_Feet_Raw)
  )
```

# Data Sources

<div style="font-size:85%;">

- Data from **Pierce County Assessor-Treasurer Data Mart**  
- Contains multiple interconnected tables  
- Data stored in `.txt` files, pipe (`|`) separated  
- Includes property, tax, sales, and valuation info  
- Each file documented with keys, types, descriptions

</div>

# Key Tables Overview

<div style="font-size:85%;">

- **Sale:** Property sales (price, date, deed type)  
- **Appraisal Account:** Land valuation, utilities  
- **Improvement:** Building features (sqft, year, stories)  
- **Improvement Built-As:** Alternate structure info  
- **Improvement Detail:** Extras (fireplaces, balconies)  
- **Tax Account:** Assessed values, exemptions  
- **Land Attribute:** Land descriptors (zoning, utilities)  
- **Seg Merge:** Parcel history (mergers, splits)  
- **Tax Description:** Parcel metadata  

</div>

# Data Merging & Cleaning
<div style="font-size: 0.8em; line-height: 1.2em;">

- Started with 250,000+ records, filtered to ~160,000 for robustness  
- Parcel Number = main join key; Building ID used for structure features  
- Steps:  
  - Joined on Parcel Number across tables  
  - Selected most recent valid sale per parcel  
  - Filtered out missing critical values (Bedrooms, Bathrooms, Stories, Year Built)  
  - Dropped variables with excessive missingness (e.g., View_Quality > 93%)  
  - Removed outliers: outside 25th–75th percentile price range and unrealistic price/sqft  
  - Included only residential, improved properties  

</div>

# Feature Construction

<div style="font-size: 0.8em; line-height: 1.2em;">

- Started with 100+ columns; applied top-down feature selection  
- Focus on interpretability, completeness, predictive relevance  
- Final features grouped conceptually into:  
  - Temporal context  
  - Size & structure  
  - Location  
  - Utilities  
  - Quality & condition  
  - Improvement  

</div>

# Data Dictionary (1/2)

<div style="font-size: 0.6em; line-height: 1.1em;">

| Variable                   | Description                       | Type    | Unit/Values                  |
|----------------------------|---------------------------------|---------|-----------------------------|
| Sale_Price_Raw*            | Final recorded sale price         | Numeric | USD                         |
| Sale_Date_Raw (TC)         | Date of sale                    | Date    | YYYY-MM-DD                  |
| Square_Feet_Raw (S)        | Total square footage            | Numeric | Square feet                 |
| Latitude_Raw (L)           | Latitude coordinate             | Numeric | Decimal degrees             |
| Longitude_Raw (L)          | Longitude coordinate            | Numeric | Decimal degrees             |
| Bedrooms_Raw (S)           | Number of bedrooms              | Numeric | Count                       |
| Bathrooms_Raw (S)          | Number of bathrooms             | Numeric | Count (including partials)  |
| Stories_Raw (S)            | Number of stories               | Numeric | Count                       |

*TC = Temporal Context, S = Size & Structure, L = Location, U = Utilities, Q = Quality & Condition, I = Improvement  
*Target variable indicated by *

</div>

# Data Dictionary (2/2)

<div style="font-size: 0.6em; line-height: 1.1em;">

| Variable                   | Description                       | Type    | Unit/Values                  |
|----------------------------|---------------------------------|---------|-----------------------------|
| Quality (Q)                | Construction quality rating       | Factor  | 11 levels                   |
| Condition (Q)              | Overall condition rating          | Factor  | 8 levels                    |
| Neighborhood (L)           | Neighborhood classification       | Factor  | ~200 levels                 |
| Street_Type (U)            | Street type classification        | Factor  | 3 levels                    |
| Utility_Water (U)          | Water utility access              | Factor  | 3 levels                    |
| Utility_Electric (U)       | Electric utility access           | Factor  | 3 levels                    |
| Utility_Sewer (U)          | Sewer utility access              | Factor  | 5 levels                    |
| Improved_Vacant_Raw (I)    | Improvement status                | Factor  | 2 levels                    |
| Year_Built_Raw (Q)         | Year constructed                 | Numeric | Year                        |
| Net_Land_Square_Feet_Raw(S)| Net land area                    | Numeric | Square feet                 |

*TC = Temporal Context, S = Size & Structure, L = Location, U = Utilities, Q = Quality & Condition, I = Improvement  

</div>

# Final Dataset Summary

<div style="font-size:0.9em; line-height:1.3;">

- 156,511 records, 18 variables  
- Residential properties in Pierce County  
- Key features: sale price & date, size, bedrooms, bathrooms, stories, year built  
- Location data: latitude, longitude, neighborhood, street type  
- Quality, condition, and utility access as factors  
- Cleaned and standardized data  
- Target: **Sale_Price_Raw**  

</div>

<div style="font-size:0.9em; line-height:1.3;">

# Target Variable: Sale Price

- Highly skewed distribution, \$0 to \$87M  
- Removed outliers:  
  - PPS > \$9,000/sqft  
  - Sale price < \$10,000  
- Cleaned range: \$10K – \$925K  
- Median: \$335,500, Mean: \$362,488  
- Distribution resembles log-normal (see boxplot)  

</div>

# Target Variable: Sale Price

```{r fig-sale-price, fig.cap="Distribution of Sales Price", fig.align="center"}
ggplot(full_data, aes(y = Sale_Price_Raw, x = "")) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar) +
  labs(
    title = "Distribution of Sales Price",
    y = "Sales Price",
    x = NULL
  ) +
  theme_minimal(base_size = 16) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

<div style="font-size:0.9em; line-height:1.3;">

# Price per Square Foot (PPS)

- Initially distorted by extreme values  
- Cleaned distribution centers around \$100–\$300/sqft  
- Critical step to reduce influence of outliers  
- PPS enables better comparison across properties  

</div>

# Price per Square Foot (PPS)

```{r fig-pps-box, fig.cap="Distribution of Price per Square Foot", fig.align="center"}
full_data <- full_data %>%
  mutate(Price_Per_SqFt = Sale_Price_Raw / Square_Feet_Raw)
ggplot(full_data, aes(x = Price_Per_SqFt)) +
  geom_boxplot() +
  labs(title = "Distribution of Price per Square Foot",
       x = "Price per SqFt") +
  theme_minimal(base_size = 15)
```

<div style="font-size: 0.9em;">
# Size & Structure

- Most homes: 1,000–4,000 sq ft  
- Price generally increases with size  
- Bedrooms: mostly 3–4; Bathrooms: 2–2.5  
- Diminishing price returns at higher counts  
- 1–2 story homes dominate; higher stories rare  
- Square footage most strongly linked to price  
- Structural variables show strong correlations → potential multicollinearity  
</div>

# Size & Structure

```{r fig-corr-ss, fig.cap="Correlation Matrix among Key Structural Variables and Sales Price", fig.align="center"}
numeric_vars <- full_data %>%
  dplyr::select(Sale_Price_Raw, Square_Feet_Raw, Bedrooms_Raw, Bathrooms_Raw, Stories_Raw) %>%
  na.omit()

cor_matrix <- cor(numeric_vars, use = "complete.obs")

corrplot(cor_matrix, method = "circle", 
         type = "upper",       
         order = "hclust",     
         addCoef.col = "black",
         tl.col = "black",     
         tl.srt = 45,          
         number.cex = 0.8,     
         diag = FALSE)         
```

<div class="slide" data-narration-layer="1" style="font-size: 0.9em;">

# Location

- Significant spatial variation observed across Pierce County  
- Heatmap of median price per sqft reveals clear clusters  
- Highest prices (~\$460/sqft) concentrated in northwestern corridor  
- Premium areas near waterfront and urban-adjacent zones  
- Lower prices in rural and southern regions  
- See figure @fig-loc-map for visual reference  

</div>

# Location

```{r fig-loc-map, fig.cap="Spatial Distribution of Median Price per Square Foot across Pierce County", fig.align="center", warning=FALSE, message=FALSE, echo=FALSE, results='hide'}
# 1. Load Pierce County boundary (Washington = '53')
pierce_county <- counties(state = "53", cb = TRUE) %>%
  filter(NAME == "Pierce")

# 2. Convert full_data to sf object with WGS84 CRS
full_data_sf <- full_data %>%
  mutate(
    Longitude_Raw = as.numeric(Longitude_Raw),
    Latitude_Raw = as.numeric(Latitude_Raw)
  ) %>%
  filter(!is.na(Longitude_Raw) & !is.na(Latitude_Raw)) %>%
  st_as_sf(coords = c("Longitude_Raw", "Latitude_Raw"), crs = 4326, remove = FALSE)

# 3. Transform Pierce County CRS to match full_data
pierce_county <- st_transform(pierce_county, st_crs(full_data_sf))

# 4. Keep only points within Pierce County
full_data_sf <- full_data_sf[st_within(full_data_sf, pierce_county, sparse = FALSE), ]

# 5. Get bounding box
lat_range <- st_bbox(pierce_county)[c("ymin", "ymax")]
lon_range <- st_bbox(pierce_county)[c("xmin", "xmax")]

# 6. Define bin edges
num_bins <- 80
lat_bins <- seq(lat_range[1], lat_range[2], length.out = num_bins + 1)
lon_bins <- seq(lon_range[1], lon_range[2], length.out = num_bins + 1)

# 7. Bin data and compute average price per sqft
full_data_sf <- full_data_sf %>%
  mutate(
    Price_per_sqft = Sale_Price_Raw / Square_Feet_Raw,
    Lat_bin = cut(Latitude_Raw, breaks = lat_bins, include.lowest = TRUE),
    Lon_bin = cut(Longitude_Raw, breaks = lon_bins, include.lowest = TRUE)
  ) %>%
  filter(Square_Feet_Raw >= 100)   # example threshold: ignore values > 1000

# 8. Aggregate by bin and extract bin boundaries
heatmap_data <- full_data_sf %>%
  filter(Square_Feet_Raw >= 100) %>%   # example threshold: ignore values > 1000
  group_by(Lat_bin, Lon_bin) %>%
  summarize(
    Avg_Price_sqft = median(Price_per_sqft, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    ymin = as.numeric(sub("\\((.+),.*", "\\1", Lat_bin)),
    ymax = as.numeric(sub(".*,([^]]*)\\]", "\\1", Lat_bin)),
    xmin = as.numeric(sub("\\((.+),.*", "\\1", Lon_bin)),
    xmax = as.numeric(sub(".*,([^]]*)\\]", "\\1", Lon_bin))
  )

# 9. Plot the heatmap using geom_rect
# Calculate log breaks evenly spaced along the log scale
log_min <- min(log(heatmap_data$Avg_Price_sqft), na.rm = TRUE)
log_max <- max(log(heatmap_data$Avg_Price_sqft), na.rm = TRUE)

# Generate 5 evenly spaced breaks on log scale
log_breaks <- seq(log_min, log_max, length.out = 5)
# Convert breaks back to original scale for labeling
legend_labels <- scales::dollar(exp(log_breaks))

ggplot() +
  geom_sf(data = pierce_county, fill = "white", color = "black", size = 0.6) +
  geom_rect(
    data = heatmap_data,
    aes(
      xmin = xmin, xmax = xmax,
      ymin = ymin, ymax = ymax,
      fill = log(Avg_Price_sqft)  # fill is log-transformed
    ),
    alpha = 0.85,
    color = NA
  ) +
  scale_fill_viridis_c(
    option = "magma",
    trans = "identity",           # no further transformation, since fill is already log
    breaks = log_breaks,          # breaks evenly spaced in log scale
    labels = legend_labels,       # labels in original scale
    na.value = "grey90",
    name = "Price per sqft"
  ) +
  coord_sf(xlim = lon_range, ylim = lat_range, expand = FALSE) +
  labs(
    title = "Median Prices of Houses",
    subtitle = "in Pierce County",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 14, hjust = 0.5, margin = margin(b = 12)),
    axis.title = element_text(face = "italic", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 9),
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 10),
    legend.key.height = unit(0.5, "cm"),  # increase legend height here
    panel.grid.major = element_line(color = "gray80", size = 0.3),
    panel.grid.minor = element_blank()
  )
```

<div style="font-size: 0.9em;">

# Neighborhood

- Median prices per sqft are right-skewed  
- Most neighborhoods: \$150–\$250/sqft  
- Top areas (e.g., 042705, 121133) exceed \$330/sqft  
- High-end areas show low price variance (tight IQR)  
- Suggests premium and stable pricing zones  
- Spatial heterogeneity important for modeling  
</div>

# Neighborhood

```{r fig-neighborhood-eda, fig.cap="Distribution of Price per Square foot in the Top and Bottom 10 Neighborhoods by Median Value", fig.align="center", warning=FALSE}
# Calculate median price per sqft by neighborhood
median_prices <- full_data_sf %>%
  group_by(Neighborhood) %>%
  summarize(median_price = median(Price_per_sqft, na.rm = TRUE)) %>%
  arrange(desc(median_price))

# Select top and bottom 10 neighborhoods
top_bottom_neigh <- bind_rows(
  head(median_prices, 10),
  tail(median_prices, 10)
)

# Filter main data to just those neighborhoods and set factor levels
filtered_data <- full_data_sf %>%
  filter(Neighborhood %in% top_bottom_neigh$Neighborhood) %>%
  mutate(Neighborhood = factor(Neighborhood, levels = top_bottom_neigh$Neighborhood))

# --- Option 1: Boxplot for Top & Bottom 10 Neighborhoods ---
ggplot(filtered_data, aes(x = Neighborhood, y = Price_per_sqft)) +
  geom_boxplot(outlier.shape = NA, fill = "#2980B9", color = "#1C2833", alpha = 0.6) +
  coord_flip(ylim = c(0, 2000)) +
  labs(
    title = "Top & Bottom 10 Neighborhoods by Median Price per Sqft",
    x = "Neighborhood",
    y = "Price per sqft"
  ) +
  theme_minimal(base_size = 12)
```

<div style="font-size: 0.9em;">
# Utilities & Access

- Missing water/electric linked to higher price per sqft  
- Likely reflects niche properties or data issues  
- Sewer access and paved streets boost price  
- Infrastructure presence = stronger value signals  
</div>

<div style="font-size: 0.9em;">
# Quality

- Sale price rises with better quality ratings  
- “Excellent” homes are clear outliers in price  
- Low/Fair homes show lower distributions  
- Strong quality-price relationship evident  
</div>

# Quality

```{r fig-price-quality, fig.cap="Relationship between Sale Price and Quality", fig.align="center", warning=FALSE, message=FALSE, echo=FALSE, results='hide'}
full_data$Quality <- factor(full_data$Quality, levels = c(
  "Low", "Low Plus", "Fair", "Fair Plus",
  "Average", "Average Plus", "Good", "Good Plus",
  "Very Good", "Very Good Plus", "Excellent"
))
ggplot(full_data, aes(x = Quality, y = Sale_Price_Raw)) +
  geom_boxplot(fill = "#3498DB", alpha = 0.6) +
  scale_y_log10(labels = scales::dollar) +
  labs(title = "Sale Price by Quality", x = "Quality", y = "Sale Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

<div style="font-size: 0.9em;">

# Condition

- Uninhabitable homes are priced significantly lower  
- Other categories overlap in price ranges  
- Condition appears weaker predictor than quality  
- Possibly more subjective or inconsistently assessed  
</div>

# Condition

```{r fig-price-cond, fig.cap="Relationship between Sale Price and Condition", fig.align="center", warning=FALSE, message=FALSE, echo=FALSE, results='hide'}
full_data$Condition <- factor(full_data$Condition, levels = c(
  "Uninhabitable", "Very Poor", "Extra Poor", "Poor", 
  "Fair", "Average", "Good", "NA"
))
ggplot(full_data, aes(x = Condition, y = Sale_Price_Raw)) +
  geom_boxplot(fill = "#27AE60", alpha = 0.6) +
  scale_y_log10(labels = scales::dollar) +
  labs(title = "Sale Price by Condition", x = "Condition", y = "Sale Price") +
  theme_minimal()
```

# Models

<div style="font-size: 0.9em;">

# Linear Model

- Started with baseline **OLS regression** for benchmarking  
- Collapsing rare **Neighborhood** levels increased error  
  - Cleaned OLS: RMSE 109,093 vs. Baseline RMSE 105,459  
- Emphasis on **MAPE** for interpretability (Baseline: 43.56%)  
- **Lasso regression** applied for regularization and variable reduction  
- Initial Lasso (with Bathrooms): RMSE 109,132; MAPE 45.72%

</div>

<div style="font-size: 0.9em;">

# Linear Model

- **Bathrooms** dropped due to multicollinearity with Bedrooms  
- Improved both OLS and Lasso performance  
- Best OLS (no Bathrooms):  
  - RMSE 99,526; MAPE 40.84%  
- Best Lasso (no Bathrooms):  
  - RMSE 99,523; MAPE **40.82%**  
- Conclusion:  
  - **MAPE minimized** with Lasso + collinearity control  
  - Compact model, high interpretability, better performance

</div>

# Linear Model

<div style="font-size: 0.7em;">

```{r tbl-model-summary, echo=FALSE}
library(knitr)

model_summary <- data.frame(
  Model = c(
    "Baseline Linear Model",
    "Cleaned Linear Model (Neighborhood Collapsed)",
    "Lasso Model",
    "OLS (No Bathrooms)",
    "Lasso (No Bathrooms)"
  ),
  MAE = c(78515.13, 81236.09, 81247.83, 77365.88, 77344.91),
  RMSE = c(105459.11, 109092.98, 109132.32, 99526.43, 99523.48),
  MAPE = c(43.56, 45.72, 45.72, 40.84, 40.82)
)

kable(model_summary, caption = "Summary of Model Performance Metrics (MAE (\\$), RMSE (\\$), MAPE (\\%))")
```

</div>

<div style="font-size: 0.9em;">

# H2O AutoML

- Applied [H2O AutoML](w) to cleaned housing dataset  
- Automatically trained multiple models (GBM, Random Forests, XGBoost, Stacked Ensembles)  
- Used **10-fold cross-validation** and **early stopping** (based on MAE)  
- Data split into training and test sets; H2O frames used  
- Up to **25 models** trained per AutoML run  
- Evaluation metrics: **MAE**, **RMSE**, **MAPE** on test set  

</div>

<div style="font-size: 0.9em;">

# H2O AutoML

- **Best model**: **Stacked Ensemble** (lowest MAE & MAPE)  
- Ensemble combines multiple base learners with a metalearner  
- Demonstrated:
  - Improved predictive accuracy over single models  
  - Robustness through stacking and automated tuning  
- Highlights AutoML’s strength in **model selection** and **ensembling**  
- Table below presents top 5 model performances on the test set  

</div>

# H2O AutoML

<div style="font-size: 0.7em;">

```{r tbl-model-summary-automl, echo=FALSE}
automl_results <- data.frame(
  Model = c(
    "GBM_1",
    "GBM_4",
    "GBM_grid_1",
    "StackedEnsemble_AllModels_1",
    "StackedEnsemble_BestOfFamily_1"
  ),
  MAE = c(31351.49, 31222.94, 31328.33, 30429.87, 31245.37),
  RMSE = c(48952.20, 49101.61, 48546.54, 47413.33, 48772.60),
  MAPE = c(10.67162, 10.53705, 10.73902, 10.25683, 10.46562)
)
automl_results <- automl_results %>%
  arrange(MAE)

kable(automl_results, caption = "AutoML Model Performance Summary (MAE, RMSE in $, MAPE in %)")
```

</div>

<div class="slide" data-narration-layer="1" style="font-size: 0.7em;">

# Neural Network

- Implemented as a fully connected feedforward neural network using **Keras**  
- Two hidden layers:
  - 128 and 32 units
  - ReLU activations
  - L2 regularization (λ = 0.01)
  - Dropout (rate = 0.1 after 1st layer)  
- Tuned hyperparameters: learning rate, batch size, regularization  
- Trained with **Adam** optimizer (lr = 0.002) using **MSE loss**  
- Metrics tracked: MAE, MAPE, and R²  
- Input data:
  - One-hot encoding for categorical features  
  - Standardization for numeric inputs  

</div>

<div class="slide" data-narration-layer="1" style="font-size: 0.7em;">

# Neural Network

- Data split: 80% training / 20% test  
- Early stopping used (patience = 100), but not triggered  
- Training stopped at **1000 epochs**  
- Test set performance:
  - **RMSE**: 54,254  
  - **MAE**: 35,744  
  - **MAPE**: 12.3%  
  - **R²**: 0.909  
- Model showed excellent generalization  
- Stable convergence, major loss drop around epoch 600  
- See plot below for training and validation loss trends

</div>

# Neural Network

```{r fig-nn-training-loss, fig.cap="Neural Network Training and Validation Loss", fig.align="center", echo=FALSE}
# load jpg file
library(jpeg)
nn_loss <- readJPEG("plots/nn_training.jpg")
ggplot() +
  annotation_custom(rasterGrob(nn_loss, width = unit(1, "npc"), height = unit(1, "npc"))) +
  labs(title = "") +
  theme_void()
```

# Cat Boost
PLACEHOLDER - not finished in PDF

<div class="slide" data-narration-layer="1" style="font-size: 0.7em;">

# XGBoost

- Used [gradient-boosted trees](w) (via **XGBoost**) for regression  
- Hyperparameter search run on 5% stratified training subset to save time  
- Dropped high-cardinality categorical variables (>100 levels)  
- Grid search over:
  - `eta`, `max_depth`, `subsample`, `colsample_bytree`, `gamma`, `num.trees`
- Evaluated with **10-fold cross-validation**  
- Target variable log-transformed for modeling  
- Final metrics computed after reversing log scale

</div>

<div class="slide" data-narration-layer="1" style="font-size: 0.7em;">

# XGBoost

- **Best configuration**:
  - `eta` = 0.25  
  - `max_depth` = 8  
  - `subsample` = 1.0  
  - `colsample_bytree` = 1.0  
  - `gamma` = 0  
  - `num.trees` = 700  
- **Cross-validated performance**:
  - **RMSE**: 592.13  
  - **MAE**: 319.16  
  - **MAPE**: 13.91%  
- XGBoost delivered robust predictions and handled complex feature interactions well

</div>

# XGBoost

```{r fig-xgboost-training-res, fig.cap="XGBoost Results from the CV", fig.align="center", echo=FALSE}
# load jpg file
library(jpeg)
nn_loss <- readJPEG("plots/xgboost_training.jpg")
ggplot() +
  annotation_custom(rasterGrob(nn_loss, width = unit(1, "npc"), height = unit(1, "npc"))) +
  labs(title = "") +
  theme_void()
```

# PLACEHOLDER ARASH MODEL

# Model Summary

<div class="slide" data-narration-layer="1" style="font-size: 0.7em;">

```{r tbl-all-summary, echo=FALSE}
library(knitr)

model_summary <- data.frame(
  Model = c(
    "Lasso Model (No Bathrooms)",
    "Neural Network",
    "AutoML StackedEnsemble_AllModels_1",
    "Tuned CatBoost Regression",
    "XGBoost Random Forest"
  ),
  MAE = c(78515.13, 35744 , 30429.87, 0, 73690.11),
  RMSE = c(105459.11, 54254, 47413.33, 0, 98453.05),
  MAPE = c(43.56, 12.3, 10.25683, 0, 28.02)
)

kable(model_summary, caption = "Summary of Model Performance Metrics (MAE (\\$), RMSE (\\$), MAPE (\\%))")

```

</div>

# Deployment

See shiny demo ...

<div class="slide" data-narration-layer="1" style="font-size: 0.8em;">

# Conclusion

- Robust approach combining data cleaning, EDA, and multiple predictive models  
- Structural features (sqft, quality), location, and utilities significantly influence prices  
- Ensemble methods (H2O AutoML Stacked Ensemble) performed best with MAPE ~10.26%  
- Shiny App integrates model for interactive prediction and exploration  
- Limitations: historical data only, no unstructured features (photos, renovations)  
- Future work: add geospatial/time trends, economic indicators, and deep learning models  

</div>