---
title: "Pierce County Property Analysis: Data Loading"
subtitle: "Step 1: Acquiring and Loading Raw Data"
author: "Your Group Name"
date: "today"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: cosmo 
    code-fold: true
    code-summary: "Show/Hide Code"
    df-print: kable # For nice table printing with knitr::kable
editor: source
execute:
  echo: true
  warning: false   # Suppress warnings globally for now
  message: false   # Suppress messages globally for now
  error: true      # Display errors if they occur
---

# 1. Introduction

This document outlines the first step in our analysis of Pierce County property data: loading all raw data files into a DuckDB in-memory database. Each text file will be loaded into its own table with column names derived from the provided metadata (PDFs). All columns will initially be loaded as `VARCHAR` (text) to ensure robust loading; type conversions will occur during the subsequent cleaning phase.

# 2. Setup and Configuration

## 2.1. Load Libraries and Functions

First, we load necessary R packages and our custom data loading function.

```{r setup-load-libs-funcs}
#| label: setup-load-libs-funcs

# Ensure these packages are installed: install.packages(c("DBI", "duckdb", "knitr", "dplyr"))
library(DBI)
library(duckdb)
library(knitr) # For kable()
library(dplyr) # For glimpse() later

# Source our custom data loading function
# The path is relative to the Quarto document's location (reports/)
source("../R/00_load_data_functions.R")

```

## 2.2. Define File Paths and Column Names

Here, we define the locations of our raw data files and the column names for each table. IMPORTANT: You MUST verify and complete the col_names_lists with the correct column names in the correct order for EACH of your files based on your metadata PDFs.

```{r}
#| label: setup-file-configs

# Base path to the directory where the raw .txt files are stored
# This path is relative to the project root, assuming the .Rproj is in the root.
# If running the .qmd directly, ensure this path correctly points to your data/raw folder.
# For Quarto rendering, it's often best to assume the project root is the working directory.
# Let's construct paths assuming the Quarto doc is in 'reports/' and data is in 'data/raw/'
# relative to a common project root.
project_root_relative_data_path <- "../data/raw/" # Path from 'reports/' dir to 'data/raw/'

file_paths_list <- list(
  sale = paste0(project_root_relative_data_path, "sale.txt"),
  appraisal_account = paste0(project_root_relative_data_path, "appraisal_account.txt"),
  improvement = paste0(project_root_relative_data_path, "improvement.txt"),
  improvement_detail = paste0(project_root_relative_data_path, "improvement_detail.txt"),
  improvement_builtas = paste0(project_root_relative_data_path, "improvement_builtas.txt"),
  land_attribute = paste0(project_root_relative_data_path, "land_attribute.txt"),
  seg_merge = paste0(project_root_relative_data_path, "seg_merge.txt"),
  tax_account = paste0(project_root_relative_data_path, "tax_account.txt"),
  tax_description = paste0(project_root_relative_data_path, "tax_description.txt")
)

# --- COLUMN NAMES ---
# !!! ACTION REQUIRED: Populate these lists accurately for EACH file !!!
# The order of names MUST match the order of columns in your .txt files.
# These are placeholders. Refer to your PDF metadata.
col_names_lists <- list(
  sale = c(
    "ETN", "Parcel_Count", "Parcel_Number", "Sale_Date_Raw", "Sale_Price_Raw", 
    "Deed_Type", "Grantor", "Grantee", "Valid_Invalid_Raw", 
    "Confirmed_Uncomfirmed_Raw", # Note: "Uncomfirmed" as per PDF
    "Exclude_Reason", "Improved_Vacant_Raw", "Appraisal_Account_Type"
  ), 
  
  appraisal_account = c(
    "Parcel_Number", "Appraisal_Account_Type", "Business_Name", "Value_Area_ID", 
    "Land_Economic_Area", "Buildings", "Group_Account_Number", 
    "Land_Gross_Acres_Raw", "Land_Net_Acres_Raw", "Land_Gross_Square_Feet_Raw", 
    "Land_Net_Square_Feet_Raw", "Land_Gross_Front_Feet_Raw", "Land_Width_Raw", 
    "Land_Depth_Raw", "Submerged_Area_Square_Feet_Raw", "Appraisal_Date_Raw", 
    "Waterfront_Type", "View_Quality", "Utility_Electric", "Utility_Sewer", 
    "Utility_Water", "Street_Type", "Latitude_Raw", "Longitude_Raw"
  ), 
  
  improvement = c(
    "Parcel_Number", "Building_ID", "Property_Type", "Neighborhood", 
    "Neighborhood_Extension", "Square_Feet_Raw", "Net_Square_Feet_Raw", 
    "Percent_Complete_Raw", "Condition", "Quality", "Primary_Occupancy_Code_Raw", 
    "Primary_Occupancy_Description", "Mobile_Home_Serial_Number", 
    "Mobile_Home_Total_Length_Raw", "Mobile_Home_Make", 
    "Attic_Finished_Square_Feet_Raw", "Basement_Square_Feet_Raw", 
    "Basement_Finished_Square_Feet_Raw", "Carport_Square_Feet_Raw", 
    "Balcony_Square_Feet_Raw", "Porch_Square_Feet_Raw", 
    "Attached_Garage_Square_Feet_Raw", "Detached_Garage_Square_Feet_Raw", 
    "Fireplaces_Raw", "Basement_Garage_Door_Raw"
  ),
                 
  improvement_detail = c(
    "Parcel_Number", "Building_ID", "Detail_Type", "Detail_Description", "Units_Raw"
  ),
                        
  improvement_builtas = c(
    "Parcel_Number", "Building_ID", "Built_As_Number_Raw", "Built_As_ID_Raw", 
    "Built_As_Description", "Built_As_Square_Feet_Raw", "HVAC_Code_Raw", 
    "HVAC_Description", "Exterior", "Interior", "Stories_Raw", "Story_Height_Raw", 
    "Sprinkler_Square_Feet_Raw", "Roof_Cover", "Bedrooms_Raw", "Bathrooms_Raw", 
    "Units_Count_Raw", "Class_Code", "Class_Description", "Year_Built_Raw", 
    "Year_Remodeled_Raw", "Adjusted_Year_Built_Raw", "Physical_Age_Raw", 
    "Built_As_Length_Raw", "Built_As_Width_Raw", "Mobile_Home_Model"
  ),
                         
  land_attribute = c(
    "Parcel_Number", "Attribute_Key", "Attribute_Description"
  ), 
                    
  seg_merge = c(
    "Seg_Merge_Number", "Parent_Child_Indicator", "Parcel_Number", 
    "Continued_Indicator", "Completed_Date_Raw", "Tax_Year_Raw"
  ), 
               
  tax_account = c(
    "Parcel_Number", "Account_Type", "Property_Type", "Site_Address", 
    "Use_Code", "Use_Description", "Tax_Year_Prior_Raw", 
    "Tax_Code_Area_Prior_Year", "Exemption_Type_Prior_Year", 
    "Current_Use_Code_Prior_Year", "Land_Value_Prior_Year_Raw", 
    "Improvement_Value_Prior_Year_Raw", "Total_Market_Value_Prior_Year_Raw", 
    "Taxable_Value_Prior_Year_Raw", "Tax_Year_Current_Raw", 
    "Tax_Code_Area_Current_Year", "Exemption_Type_Current_Year", 
    "Current_Use_Code_Current_Year", "Land_Value_Current_Year_Raw", 
    "Improvement_Value_Current_Year_Raw", "Total_Market_Value_Current_Year_Raw", 
    "Taxable_Value_Current_Year_Raw", "Range", "Township", "Section", 
    "Quarter_Section", "Subdivision_Name", "Located_On_Parcel"
  ), 
                 
  tax_description = c(
    "Parcel_Number", "Line_Number_Raw", "Tax_Description_Line"
  )
)

# Verify all files have column name definitions (basic check)
if (!all(names(file_paths_list) %in% names(col_names_lists))) {
  stop("Mismatch between file_paths_list and col_names_lists. Ensure every file has a corresponding column name definition.")
}
if (!all(names(col_names_lists) %in% names(file_paths_list))) {
  stop("Mismatch between col_names_lists and file_paths_list. Ensure every column name definition has a corresponding file.")
}
```

## 2.3. Initialize DuckDB Connection

We'll use an in-memory DuckDB database for this session.

```{r}
#| label: setup-duckdb-connection

con <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
cat("DuckDB in-memory connection established.\n")
```

# 3. Load Raw Data into DuckDB

Now, we iterate through our configured files and load each one into a separate table in DuckDB using our custom function.

```{r}
#| label: load-data-to-duckdb
#| results: 'asis' # Allows cat() HTML output to render directly

cat("### Starting Data Loading Process:\n\n")

loaded_successfully <- c() # To track which tables loaded

for (table_key in names(file_paths_list)) {
  file_path <- file_paths_list[[table_key]]
  col_names <- col_names_lists[[table_key]]
  target_table_name <- paste0(table_key, "_raw_duckdb") # e.g., "sale_raw_duckdb"
  
  cat(paste0("Attempting to load: ", basename(file_path), " into table `", target_table_name, "`...\n"))
  
  if (is.null(col_names) || length(col_names) == 0) {
      cat(paste0("<p style='color:red;'><strong>Error:</strong> Column names for '", table_key, "' are not defined or empty. Skipping.</p>\n\n"))
      loaded_successfully[target_table_name] <- FALSE
      next
  }
  
  success <- load_pipe_delimited_file_to_duckdb(
    con = con,
    file_path = file_path,
    col_names_vector = col_names,
    target_table_name = target_table_name
  )
  loaded_successfully[target_table_name] <- success
}

cat("\n### Data Loading Process Complete.\n")

# Summary of loading
cat("\n#### Loading Summary:\n")
for(tbl_name in names(loaded_successfully)){
    status_msg <- if(loaded_successfully[tbl_name]) "Successfully loaded" else "Failed to load or file not found"
    cat(paste0("- `", tbl_name, "`: ", status_msg, "\n"))
}
```

# 4. Preliminary Data Inspection

Let's list the tables in our DuckDB database and look at the first few rows and structure of each loaded table to verify the loading process.

```{r}
#| label: inspect-loaded-tables
#| results: 'asis'

cat("\n### Tables in DuckDB:\n")
loaded_tables_in_db <- DBI::dbListTables(con)
if (length(loaded_tables_in_db) > 0) {
  kable(loaded_tables_in_db, col.names = "Table Name", caption = "Tables present in DuckDB") %>% print()
} else {
  cat("No tables found in DuckDB. Please check loading logs.\n")
}


cat("\n\n### Preview of Loaded Tables (First 3 Rows and Structure):\n")
for (table_name_raw in loaded_tables_in_db) {
  # Only preview tables we attempted to load based on our naming convention
  if (grepl("_raw_duckdb$", table_name_raw) && isTRUE(loaded_successfully[table_name_raw])) {
    cat(paste0("\n#### Table: `", table_name_raw, "`\n"))
    
    # Get column count from DB
    db_cols <- DBI::dbListFields(con, table_name_raw)
    num_db_cols <- length(db_cols)
    
    # Get defined column count
    original_key <- sub("_raw_duckdb$", "", table_name_raw) # e.g. "sale" from "sale_raw_duckdb"
    num_defined_cols <- length(col_names_lists[[original_key]])

    cat(paste0("*Defined columns: ", num_defined_cols, ", Columns in DB table: ", num_db_cols, "*\n"))
    if (num_defined_cols != num_db_cols && num_db_cols > 0) {
         cat(paste0("<p style='color:orange;'><strong>Warning:</strong> Mismatch in column count for `", table_name_raw, 
                    "`. Defined: ", num_defined_cols, ", Actual in DB: ", num_db_cols, 
                    ". This could indicate issues with delimiter, quoting, or column name definitions.</p>"))
    }

    # Preview first 3 rows
    cat("\n##### First 3 Rows:\n")
    tryCatch({
      preview_data <- DBI::dbGetQuery(con, paste0("SELECT * FROM ", table_name_raw, " LIMIT 3"))
      if (nrow(preview_data) > 0) {
        kable(preview_data, caption = paste("First 3 rows of", table_name_raw)) %>% 
          kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive"), 
                                    full_width = FALSE,
                                    font_size = 10) %>% # Smaller font for wide tables
          print()
      } else {
        cat("Table is empty or could not retrieve rows.\n")
      }
    }, error = function(e) {
      cat(paste0("<p style='color:red;'>Error previewing table `", table_name_raw, "`: ", e$message, "</p>\n"))
    })
    
    # Show structure (column names and types from DuckDB's perspective)
    cat("\n##### Structure (from DuckDB):\n")
    tryCatch({
        # DuckDB's PRAGMA table_info('table_name') is good for this
        structure_info <- DBI::dbGetQuery(con, paste0("PRAGMA table_info('", table_name_raw, "');"))
        if (nrow(structure_info) > 0) {
            kable(structure_info %>% select(name, type), caption = paste("Structure of", table_name_raw)) %>% 
              kableExtra::kable_styling(bootstrap_options = c("condensed"), full_width = FALSE) %>%
              print()
        } else {
            cat("Could not retrieve structure information.\n")
        }
    }, error = function(e) {
      cat(paste0("<p style='color:red;'>Error getting structure for table `", table_name_raw, "`: ", e$message, "</p>\n"))
    })
    cat("\n---\n") # Separator
  }
}
```

# 5. Next Steps

With the raw data loaded, the next phase will involve:

1.  Detailed Data Cleaning: For each table, convert data types, handle missing values, trim whitespace, and correct inconsistencies. This will be done using SQL queries in DuckDB, potentially defined in separate R functions for modularity.

2.  Data Integration: Joining these cleaned tables into one or more master analytical datasets based on the relationships defined in the datamart_diagram.pdf.

# 6. Close DuckDB Connection (End of Session)

```{r}
#| label: cleanup-duckdb-connection
#| echo: false 
#| include: false 

if (exists("con") && DBI::dbIsValid(con)) {
  DBI::dbDisconnect(con, shutdown = TRUE)
  cat("DuckDB connection closed.\n") # Message suppressed by include=false
}
rm(list = ls()) # Clean up environment


```
