---
title: "Basic Modelling"
subtitle: "Classification"
author: "Your Group Name"
date: "today"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: cosmo 
    code-fold: true
    code-summary: "Show/Hide Code"
    df-print: kable # For nice table printing with knitr::kable
editor: source
execute:
  echo: true
  warning: false   # Suppress warnings globally for now
  message: false   # Suppress messages globally for now
  error: true      # Display errors if they occur
---

# Datagather

# 1. Introduction

This document outlines the first step in our analysis of Pierce County property data: loading all raw data files into a DuckDB in-memory database. Each text file will be loaded into its own table with column names derived from the provided metadata (PDFs). All columns will initially be loaded as `VARCHAR` (text) to ensure robust loading; type conversions will occur during the subsequent cleaning phase.

# 2. Setup and Configuration

## 2.1. Load Libraries and Functions

First, we load necessary R packages and our custom data loading function.

```{r setup-load-libs-funcs}
#| code-summary: Libraries
#| code-fold: true
#| code-hidden: true
#| label: setup-load-libs-funcs

if(!require(DBI)) install.packages("DBI");
if(!require(duckdb)) install.packages("duckdb");
if(!require(knitr)) install.packages("knitr");
if(!require(dplyr)) install.packages("dplyr");
library(DBI)
library(duckdb)
library(knitr)
library(dplyr)

# Source our custom data loading function
# The path is relative to the Quarto document's location (reports/)
source("../R/00_load_data_functions.R")

```

## 2.2. Define File Paths and Column Names

Here, we define the locations of our raw data files and the column names for each table. IMPORTANT: You MUST verify and complete the col_names_lists with the correct column names in the correct order for EACH of your files based on your metadata PDFs.

```{r}
#| label: setup-file-configs

# Base path to the directory where the raw .txt files are stored
# This path is relative to the project root, assuming the .Rproj is in the root.
# If running the .qmd directly, ensure this path correctly points to your data/raw folder.
# For Quarto rendering, it's often best to assume the project root is the working directory.
# Let's construct paths assuming the Quarto doc is in 'reports/' and data is in 'data/raw/'
# relative to a common project root.
project_root_relative_data_path <- "../data/raw/" # Path from 'reports/' dir to 'data/raw/'

file_paths_list <- list(
  sale = paste0(project_root_relative_data_path, "sale.txt"),
  appraisal_account = paste0(project_root_relative_data_path, "appraisal_account.txt"),
  improvement = paste0(project_root_relative_data_path, "improvement.txt"),
  improvement_detail = paste0(project_root_relative_data_path, "improvement_detail.txt"),
  improvement_builtas = paste0(project_root_relative_data_path, "improvement_builtas.txt"),
  land_attribute = paste0(project_root_relative_data_path, "land_attribute.txt"),
  seg_merge = paste0(project_root_relative_data_path, "seg_merge.txt"),
  tax_account = paste0(project_root_relative_data_path, "tax_account.txt"),
  tax_description = paste0(project_root_relative_data_path, "tax_description.txt")
)

# --- COLUMN NAMES ---
# !!! ACTION REQUIRED: Populate these lists accurately for EACH file !!!
# The order of names MUST match the order of columns in your .txt files.
# These are placeholders. Refer to your PDF metadata.
col_names_lists <- list(
  sale = c(
    "ETN", "Parcel_Count", "Parcel_Number", "Sale_Date_Raw", "Sale_Price_Raw", 
    "Deed_Type", "Grantor", "Grantee", "Valid_Invalid_Raw", 
    "Confirmed_Uncomfirmed_Raw", # Note: "Uncomfirmed" as per PDF
    "Exclude_Reason", "Improved_Vacant_Raw", "Appraisal_Account_Type"
  ), 
  
  appraisal_account = c(
    "Parcel_Number", "Appraisal_Account_Type", "Business_Name", "Value_Area_ID", 
    "Land_Economic_Area", "Buildings", "Group_Account_Number", 
    "Land_Gross_Acres_Raw", "Land_Net_Acres_Raw", "Land_Gross_Square_Feet_Raw", 
    "Land_Net_Square_Feet_Raw", "Land_Gross_Front_Feet_Raw", "Land_Width_Raw", 
    "Land_Depth_Raw", "Submerged_Area_Square_Feet_Raw", "Appraisal_Date_Raw", 
    "Waterfront_Type", "View_Quality", "Utility_Electric", "Utility_Sewer", 
    "Utility_Water", "Street_Type", "Latitude_Raw", "Longitude_Raw"
  ), 
  
  improvement = c(
    "Parcel_Number", "Building_ID", "Property_Type", "Neighborhood", 
    "Neighborhood_Extension", "Square_Feet_Raw", "Net_Square_Feet_Raw", 
    "Percent_Complete_Raw", "Condition", "Quality", "Primary_Occupancy_Code_Raw", 
    "Primary_Occupancy_Description", "Mobile_Home_Serial_Number", 
    "Mobile_Home_Total_Length_Raw", "Mobile_Home_Make", 
    "Attic_Finished_Square_Feet_Raw", "Basement_Square_Feet_Raw", 
    "Basement_Finished_Square_Feet_Raw", "Carport_Square_Feet_Raw", 
    "Balcony_Square_Feet_Raw", "Porch_Square_Feet_Raw", 
    "Attached_Garage_Square_Feet_Raw", "Detached_Garage_Square_Feet_Raw", 
    "Fireplaces_Raw", "Basement_Garage_Door_Raw"
  ),
                 
  improvement_detail = c(
    "Parcel_Number", "Building_ID", "Detail_Type", "Detail_Description", "Units_Raw"
  ),
                        
  improvement_builtas = c(
    "Parcel_Number", "Building_ID", "Built_As_Number_Raw", "Built_As_ID_Raw", 
    "Built_As_Description", "Built_As_Square_Feet_Raw", "HVAC_Code_Raw", 
    "HVAC_Description", "Exterior", "Interior", "Stories_Raw", "Story_Height_Raw", 
    "Sprinkler_Square_Feet_Raw", "Roof_Cover", "Bedrooms_Raw", "Bathrooms_Raw", 
    "Units_Count_Raw", "Class_Code", "Class_Description", "Year_Built_Raw", 
    "Year_Remodeled_Raw", "Adjusted_Year_Built_Raw", "Physical_Age_Raw", 
    "Built_As_Length_Raw", "Built_As_Width_Raw", "Mobile_Home_Model"
  ),
                         
  land_attribute = c(
    "Parcel_Number", "Attribute_Key", "Attribute_Description"
  ), 
                    
  seg_merge = c(
    "Seg_Merge_Number", "Parent_Child_Indicator", "Parcel_Number", 
    "Continued_Indicator", "Completed_Date_Raw", "Tax_Year_Raw"
  ), 
               
  tax_account = c(
    "Parcel_Number", "Account_Type", "Property_Type", "Site_Address", 
    "Use_Code", "Use_Description", "Tax_Year_Prior_Raw", 
    "Tax_Code_Area_Prior_Year", "Exemption_Type_Prior_Year", 
    "Current_Use_Code_Prior_Year", "Land_Value_Prior_Year_Raw", 
    "Improvement_Value_Prior_Year_Raw", "Total_Market_Value_Prior_Year_Raw", 
    "Taxable_Value_Prior_Year_Raw", "Tax_Year_Current_Raw", 
    "Tax_Code_Area_Current_Year", "Exemption_Type_Current_Year", 
    "Current_Use_Code_Current_Year", "Land_Value_Current_Year_Raw", 
    "Improvement_Value_Current_Year_Raw", "Total_Market_Value_Current_Year_Raw", 
    "Taxable_Value_Current_Year_Raw", "Range", "Township", "Section", 
    "Quarter_Section", "Subdivision_Name", "Located_On_Parcel"
  ), 
                 
  tax_description = c(
    "Parcel_Number", "Line_Number_Raw", "Tax_Description_Line"
  )
)

# Verify all files have column name definitions (basic check)
if (!all(names(file_paths_list) %in% names(col_names_lists))) {
  stop("Mismatch between file_paths_list and col_names_lists. Ensure every file has a corresponding column name definition.")
}
if (!all(names(col_names_lists) %in% names(file_paths_list))) {
  stop("Mismatch between col_names_lists and file_paths_list. Ensure every column name definition has a corresponding file.")
}
```

## 2.3. Initialize DuckDB Connection

We'll use an in-memory DuckDB database for this session.

```{r}
#| label: setup-duckdb-connection

con <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
cat("DuckDB in-memory connection established.\n")
```

# 3. Load Raw Data into DuckDB

Now, we iterate through our configured files and load each one into a separate table in DuckDB using our custom function.

```{r}
#| label: load-data-to-duckdb
#| results: 'asis' # Allows cat() HTML output to render directly

cat("### Starting Data Loading Process:\n\n")

loaded_successfully <- c() # To track which tables loaded

for (table_key in names(file_paths_list)) {
  file_path <- file_paths_list[[table_key]]
  col_names <- col_names_lists[[table_key]]
  target_table_name <- paste0(table_key, "_raw_duckdb") # e.g., "sale_raw_duckdb"
  
  cat(paste0("Attempting to load: ", basename(file_path), " into table `", target_table_name, "`...\n"))
  
  if (is.null(col_names) || length(col_names) == 0) {
      cat(paste0("<p style='color:red;'><strong>Error:</strong> Column names for '", table_key, "' are not defined or empty. Skipping.</p>\n\n"))
      loaded_successfully[target_table_name] <- FALSE
      next
  }
  
  success <- load_pipe_delimited_file_to_duckdb(
    con = con,
    file_path = file_path,
    col_names_vector = col_names,
    target_table_name = target_table_name
  )
  loaded_successfully[target_table_name] <- success
}

cat("\n### Data Loading Process Complete.\n")

# Summary of loading
cat("\n#### Loading Summary:\n")
for(tbl_name in names(loaded_successfully)){
    status_msg <- if(loaded_successfully[tbl_name]) "Successfully loaded" else "Failed to load or file not found"
    cat(paste0("- `", tbl_name, "`: ", status_msg, "\n"))
}
```

# 4. Preliminary Data Inspection

Let's list the tables in our DuckDB database and look at the first few rows and structure of each loaded table to verify the loading process.

```{r}
#| label: inspect-loaded-tables
#| results: 'asis'

cat("\n### Tables in DuckDB:\n")
loaded_tables_in_db <- DBI::dbListTables(con)
if (length(loaded_tables_in_db) > 0) {
  kable(loaded_tables_in_db, col.names = "Table Name", caption = "Tables present in DuckDB") %>% print()
} else {
  cat("No tables found in DuckDB. Please check loading logs.\n")
}


cat("\n\n### Preview of Loaded Tables (First 3 Rows and Structure):\n")
for (table_name_raw in loaded_tables_in_db) {
  # Only preview tables we attempted to load based on our naming convention
  if (grepl("_raw_duckdb$", table_name_raw) && isTRUE(loaded_successfully[table_name_raw])) {
    cat(paste0("\n#### Table: `", table_name_raw, "`\n"))
    
    # Get column count from DB
    db_cols <- DBI::dbListFields(con, table_name_raw)
    num_db_cols <- length(db_cols)
    
    # Get defined column count
    original_key <- sub("_raw_duckdb$", "", table_name_raw) # e.g. "sale" from "sale_raw_duckdb"
    num_defined_cols <- length(col_names_lists[[original_key]])

    cat(paste0("*Defined columns: ", num_defined_cols, ", Columns in DB table: ", num_db_cols, "*\n"))
    if (num_defined_cols != num_db_cols && num_db_cols > 0) {
         cat(paste0("<p style='color:orange;'><strong>Warning:</strong> Mismatch in column count for `", table_name_raw, 
                    "`. Defined: ", num_defined_cols, ", Actual in DB: ", num_db_cols, 
                    ". This could indicate issues with delimiter, quoting, or column name definitions.</p>"))
    }

    # Preview first 3 rows
    cat("\n##### First 3 Rows:\n")
    tryCatch({
      preview_data <- DBI::dbGetQuery(con, paste0("SELECT * FROM ", table_name_raw, " LIMIT 3"))
      if (nrow(preview_data) > 0) {
        kable(preview_data, caption = paste("First 3 rows of", table_name_raw)) %>% 
          kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive"), 
                                    full_width = FALSE,
                                    font_size = 10) %>% # Smaller font for wide tables
          print()
      } else {
        cat("Table is empty or could not retrieve rows.\n")
      }
    }, error = function(e) {
      cat(paste0("<p style='color:red;'>Error previewing table `", table_name_raw, "`: ", e$message, "</p>\n"))
    })
    
    # Show structure (column names and types from DuckDB's perspective)
    cat("\n##### Structure (from DuckDB):\n")
    tryCatch({
        # DuckDB's PRAGMA table_info('table_name') is good for this
        structure_info <- DBI::dbGetQuery(con, paste0("PRAGMA table_info('", table_name_raw, "');"))
        if (nrow(structure_info) > 0) {
            kable(structure_info %>% select(name, type), caption = paste("Structure of", table_name_raw)) %>% 
              kableExtra::kable_styling(bootstrap_options = c("condensed"), full_width = FALSE) %>%
              print()
        } else {
            cat("Could not retrieve structure information.\n")
        }
    }, error = function(e) {
      cat(paste0("<p style='color:red;'>Error getting structure for table `", table_name_raw, "`: ", e$message, "</p>\n"))
    })
    cat("\n---\n") # Separator
  }
}
```

# 5. Modeling Preparation
```{r}
# select 80k Appraisal_Account records and join the latest Sale record
# There appears some Error with empty values in
# appraisal_account_raw_duckdb (relevant) -- it works however with the first 80k rows
# tax_account_raw_duckdb (irrelevant)
# tax_description_raw_duckdb (irrelevant)

# get list of tables
DBI::dbListTables(con)

# retrieve apraisal_account_raw_duckdb
appraisal_account <- DBI::dbGetQuery(con, "SELECT * FROM appraisal_account_raw_duckdb limit 80000")

# retrieve sale_raw_duckdb and join on Parcel_Number
sales <- DBI::dbGetQuery(con, "SELECT * FROM sale_raw_duckdb")
sales <- appraisal_account %>%
  dplyr::inner_join(sales, by = "Parcel_Number") 

# get the most recent sale per parcel
sales <- sales %>%
  dplyr::group_by(Parcel_Number) %>%
  dplyr::slice_max(Sale_Date_Raw, n = 1) %>%
  dplyr::ungroup()

# drop Parcels that appear more than once
sales <- sales %>%
  dplyr::group_by(Parcel_Number) %>%
  dplyr::mutate(Count = n()) %>%
  dplyr::filter(Count == 1) %>%
  dplyr::ungroup() %>%
  dplyr::select(-Count)

# print unique parcels
unique_parcels <- unique(sales$Parcel_Number)
cat(paste0("Unique parcels in the most recent sales: ", length(unique_parcels), "\n"))

```

```{r}
# select improvement_builtas_raw_duckdb and improvement_raw_duckdb and join on Parcel_Number add Building_ID

improvement <- DBI::dbGetQuery(con, "SELECT * FROM improvement_raw_duckdb")
improvement_builtas <- DBI::dbGetQuery(con, "SELECT * FROM improvement_builtas_raw_duckdb")
improvement <- improvement %>%
  dplyr::inner_join(improvement_builtas, by = c("Parcel_Number", "Building_ID"))
improvement <- improvement %>%
  dplyr::group_by(Parcel_Number) %>%
  dplyr::slice_max(Year_Built_Raw, n = 1) %>%
  dplyr::ungroup()
```

```{r}
# join improvement and sales on Parcel_Number
# filter for Residential, Valid, Improved
# TODO: currently double Parcel_Number are dropped. check if this is correct (mutliple improvements). but must have just one anyway
# 21k Rows left for EDA

full_data <- sales %>%
  dplyr::inner_join(improvement, by = "Parcel_Number") %>%
  dplyr::filter(Appraisal_Account_Type.x == "Residential") %>%
  dplyr::filter(Valid_Invalid_Raw == "1") %>%
  dplyr::filter(Improved_Vacant_Raw == "1")

#drop Parcels that appear more than once
full_data <- full_data %>%
  dplyr::group_by(Parcel_Number) %>%
  dplyr::mutate(Count = n()) %>%
  dplyr::filter(Count == 1) %>%
  dplyr::ungroup() %>%
  dplyr::select(-Count)

# print unique parcels
unique_parcels <- unique(full_data$Parcel_Number)
cat(paste0("Unique parcels in the most recent sales: ", length(unique_parcels), "\n"))

full_data <- full_data %>%
  dplyr::select(Parcel_Number, Sale_Date_Raw, Sale_Price_Raw, Year_Built_Raw, Bedrooms_Raw, Bathrooms_Raw, Square_Feet_Raw, 
                Latitude_Raw, Longitude_Raw, Neighborhood, View_Quality, Stories_Raw, Adjusted_Year_Built_Raw, Quality, Condition, 
                Utility_Electric, Utility_Sewer, Utility_Water, Street_Type, Valid_Invalid_Raw, Improved_Vacant_Raw)
```

```{r}
# Convert to the right data types
# this could be stored in DB, csv etc. and loaded in the future

full_data <- full_data %>%
  mutate(
    Sale_Date_Raw = as.Date(Sale_Date_Raw, format = "%m/%d/%Y"),
    Sale_Price_Raw = as.numeric(Sale_Price_Raw),
    Year_Built_Raw = as.numeric(Year_Built_Raw),
    Adjusted_Year_Built_Raw = as.numeric(Adjusted_Year_Built_Raw),
    Bedrooms_Raw = as.numeric(Bedrooms_Raw),
    Bathrooms_Raw = as.numeric(Bathrooms_Raw),
    Square_Feet_Raw = as.numeric(Square_Feet_Raw),
    Latitude_Raw = as.numeric(Latitude_Raw),
     Stories_Raw = as.numeric(Stories_Raw), 
    Longitude_Raw = as.numeric(Longitude_Raw),
    Neighborhood = as.factor(Neighborhood),
    View_Quality = as.factor(View_Quality),
    Quality = as.factor(Quality),
    Condition = as.factor(Condition),
    Utility_Electric = as.factor(Utility_Electric),
    Utility_Sewer = as.factor(Utility_Sewer),
    Utility_Water = as.factor(Utility_Water),
    Street_Type = as.factor(Street_Type),
    Valid_Invalid_Raw = as.factor(Valid_Invalid_Raw),
    Improved_Vacant_Raw = as.factor(Improved_Vacant_Raw)
  )
```

# Model

```{r}
#| code-summary: Libraries
#| code-fold: true
#| code-hidden: true
#| label: setup-load-libs-funcs

# Only numeric, non-NA data for PCA
if(!require(tidyr)) install.packages("tidyr");
if(!require(ggplot2)) install.packages("ggplot2");
if(!require(FactoMineR)) install.packages("FactoMineR");
if(!require(factoextra)) install.packages("factoextra");
library(tidyr)
library(ggplot2)
library(FactoMineR)
library(factoextra)

# Group raw variables into driving factors
size_vars <- c("Square_Feet_Raw", "Bedrooms_Raw", "Bathrooms_Raw", "Stories_Raw")
location_vars <- c("Latitude_Raw", "Longitude_Raw", "Neighborhood", "Street_Type")
quality_vars <- c("Year_Built_Raw", "Adjusted_Year_Built_Raw", "View_Quality", 
                  "Condition", "Quality", "Utility_Electric", "Utility_Sewer", "Utility_Water")


# Select numeric columns from full_data
numeric_data <- full_data %>%
  select(Sale_Price_Raw, all_of(size_vars), all_of(location_vars[1:2]), all_of(quality_vars[1:2])) %>%
  drop_na()

# Scale data
scaled_data <- scale(numeric_data)

# run PCA
pca_result <- prcomp(scaled_data, center = TRUE, scale. = TRUE)

# scree plot
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50))

# contribution of variables to first components
fviz_pca_var(pca_result, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)

```

```{r}
final_model_data <- full_data %>%
  select(Sale_Price_Raw, all_of(size_vars), all_of(location_vars), all_of(quality_vars))
final_model_data
```

```{r}
# list unique values, mins, maxs, means, medians, how much NA, etc.
summarize_final_model_data <- function(df) {
  tibble(
    Variable = names(df),
    Type = sapply(df, function(x) class(x)[1]),
    Unique_Values = sapply(df, function(x) n_distinct(x, na.rm = TRUE)),
    NAs = sapply(df, function(x) sum(is.na(x))),
    Min = sapply(df, function(x) if(is.numeric(x)) min(x, na.rm = TRUE) else NA),
    Max = sapply(df, function(x) if(is.numeric(x)) max(x, na.rm = TRUE) else NA),
    Mean = sapply(df, function(x) if(is.numeric(x)) mean(x, na.rm = TRUE) else NA),
    Median = sapply(df, function(x) if(is.numeric(x)) median(x, na.rm = TRUE) else NA)
  )
}

summary_stats <- summarize_final_model_data(final_model_data)
print(summary_stats)
```

```{r}
# drop View_quality
final_model_data <- final_model_data %>%
  select(-View_Quality)
```

```{r}
missing_summary <- final_model_data %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Percent") %>%
  arrange(desc(Missing_Percent))
print(missing_summary)
```

## Bedroom imputation
```{r}
final_model_data %>%
  mutate(Missing_Bedrooms = is.na(Bedrooms_Raw)) %>%
  group_by(Missing_Bedrooms) %>%
  summarise(
    Mean_Price = mean(Sale_Price_Raw, na.rm = TRUE),
    Mean_Size = mean(Square_Feet_Raw, na.rm = TRUE),
    Mean_Bathrooms = mean(Bathrooms_Raw, na.rm = TRUE),
    Count = n()
  )
```

```{r}
# impute missing values for bedrooms with KNN
if(!require(VIM)) install.packages("VIM");
library(VIM)


# KNN imputation
knn_data <- final_model_data %>%
  select(Sale_Price_Raw, Square_Feet_Raw, Latitude_Raw, Longitude_Raw, 
         Bedrooms_Raw, Bathrooms_Raw, Stories_Raw)  # only numeric

imputed_data <- kNN(knn_data, variable = c("Bedrooms_Raw", "Bathrooms_Raw", "Stories_Raw"), k = 5)

imputed_data_clean <- imputed_data %>%
  select(-ends_with(".imp"))
```

```{r}
table(is.na(final_model_data$Bedrooms_Raw))
table(is.na(imputed_data_clean$Bedrooms_Raw))
```

## Bathroom imputation
```{r}
final_model_data %>%
  mutate(Missing_Bathrooms = is.na(Bathrooms_Raw)) %>%
  group_by(Missing_Bathrooms) %>%
  summarise(
    Mean_Price = mean(Sale_Price_Raw, na.rm = TRUE),
    Mean_Size = mean(Square_Feet_Raw, na.rm = TRUE),
    Mean_Bedrooms = mean(Bedrooms_Raw, na.rm = TRUE),
    Count = n()
  )
```

```{r}
# impute missing values for bedrooms with KNN
if(!require(VIM)) install.packages("VIM");
library(VIM)


# KNN imputation
knn_data <- final_model_data %>%
  select(Sale_Price_Raw, Square_Feet_Raw, Latitude_Raw, Longitude_Raw, 
         Bedrooms_Raw, Bathrooms_Raw, Stories_Raw)  # only numeric

imputed_data <- kNN(knn_data, variable = c("Bedrooms_Raw", "Bathrooms_Raw", "Stories_Raw"), k = 5)

imputed_data_clean <- imputed_data %>%
  select(-ends_with(".imp"))
```

```{r}
table(is.na(final_model_data$Bathrooms_Raw))
table(is.na(imputed_data_clean$Bathrooms_Raw))
```

## Stories imputation
```{r}
final_model_data %>%
  mutate(Missing_stories = is.na(Stories_Raw)) %>%
  group_by(Missing_stories) %>%
  summarise(
    Mean_Price = mean(Sale_Price_Raw, na.rm = TRUE),
    Mean_Size = mean(Square_Feet_Raw, na.rm = TRUE),
    Mean_Bedrooms = mean(Bedrooms_Raw, na.rm = TRUE),
    Mean_Bathrooms = mean(Bathrooms_Raw, na.rm = TRUE),
    Count = n()
  )
```

```{r}
# impute missing values for bedrooms with KNN
if(!require(VIM)) install.packages("VIM");
library(VIM)


# KNN imputation
knn_data <- final_model_data %>%
  select(Sale_Price_Raw, Square_Feet_Raw, Latitude_Raw, Longitude_Raw, 
         Bedrooms_Raw, Bathrooms_Raw, Stories_Raw)  # only numeric

imputed_data <- kNN(knn_data, variable = c("Bedrooms_Raw", "Bathrooms_Raw", "Stories_Raw"), k = 5)

imputed_data_clean <- imputed_data %>%
  select(-ends_with(".imp"))
```

```{r}
table(is.na(final_model_data$Stories_Raw))
table(is.na(imputed_data_clean$Stories_Raw))
```

```{r}
final_model_data <- final_model_data %>%
  mutate(
    Bedrooms_Raw = imputed_data_clean$Bedrooms_Raw,
    Bathrooms_Raw = imputed_data_clean$Bathrooms_Raw,
    Stories_Raw = imputed_data_clean$Stories_Raw
  )
```


```{r}
#| code-summary: Libraries
#| code-fold: true

if(!require(caret)) install.packages("caret");
if(!require(randomForest)) install.packages("randomForest");
if(!require(e1071)) install.packages("e1071");

library(caret)
library(randomForest)
library(e1071)

# split data into training and test set
train_index <- createDataPartition(final_model_data$Sale_Price_Raw, p = 0.8, list = FALSE)
train_data <- final_model_data[train_index, ]
test_data <- final_model_data[-train_index, ]

train_data <- train_data %>% select(-Neighborhood)
test_data <- test_data %>% select(-Neighborhood)

# train model
ntree_vals <- seq(50, 1000, by = 50)
ctrl <- trainControl(method = "cv", number = 5)
rf_models <- lapply(ntree_vals, function(ntree) {
  train(x = train_data,
        y = train_data$Sale_Price_Raw,
        method = "rf",
        ntree = ntree,
        trControl = ctrl,
        tuneGrid = expand.grid(mtry = floor(sqrt(ncol(train_data)))),
        )
})

rf_results <- sapply(rf_models, function(model) {
  model$results$Accuracy
})

rf_results <- data.frame(ntree = ntree_vals, Accuracy = rf_results)
best <- rf_results[which.max(rf_results$Accuracy), ]
print(best)
```

```{r}
library(ggplot2)

ggplot(data = data.frame(Actual = test_data$Sale_Price_Raw,
                         Predicted = predictions),
       aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Predicted vs. Actual Sale Prices",
       x = "Actual Sale Price", y = "Predicted Sale Price") +
  theme_minimal()
```




