---
title: "Random Forest"
author: "Hakim Bajim"
format: pdf
editor: visual
---

# Packages
```{r}
if(!require(tidyr)) install.packages("tidyr"); library(tidyr);
if(!require(ggplot2)) install.packages("ggplot2"); library(ggplot2);
if(!require(FactoMineR)) install.packages("FactoMineR"); library(FactoMineR);
if(!require(factoextra)) install.packages("factoextra"); library(factoextra);
if(!require(cluster)) install.packages("cluster"); library(cluster);
if(!require(dplyr)) install.packages("dplyr"); library(dplyr);
if(!require(ggdendro)) install.packages("ggdendro"); library(ggdendro);
if(!require(ranger)) install.packages("ranger"); library(ranger);
if(!require(purrr)) install.packages("purrr"); library(purrr);
if(!require(progress)) install.packages("progress"); library(progress);
if(!require(xgboost)) install.packages("xgboost"); library(xgboost);
```

# Data loading
```{r}
data = read.csv("../data/final_data.csv")
head(data)
```

```{r}
# get columns names
colnames(data)
```
```{r}
# describe the data
summary(data)
```

# Random Forest Model
## Preperation
```{r}
set.seed(42)

# convert char columns to factors
data <- data %>%
  mutate(across(where(is.character), as.factor))

data <- data %>%
  mutate(Sale_Price_log = log(Sale_Price_Raw + 1))

n <- nrow(data)
train_indices <- sample(seq_len(n), size = 0.8 * n)

train_data <- data[train_indices, ]
test_data  <- data[-train_indices, ]

# create 10 folds
n <- nrow(train_data)
folds <- sample(rep(1:10, length.out = n))

# hyperparams
trees_to_test <- c(200, 300, 500, 700)
```

## Random Forest with Ranger
```{r}
cv_results <- data.frame(num.trees = integer(), fold = integer(), RMSE = double(), MAE = double(), MAPE = double())

# perform CV with batching
for (ntree in trees_to_test) {
  cat("Evaluating num.trees =", ntree, "\n")
  pb <- progress_bar$new(format = "(:spin) [:bar] :percent [Elapsed time: :elapsedfull || Estimated time remaining: :eta]",
                       total = 10,
                       complete = "=",   # Completion bar character
                       incomplete = "-", # Incomplete bar character
                       current = ">",    # Current bar character
                       clear = FALSE,    # If TRUE, clears the bar when finish
                       width = 100)      # Width of the progress bar
  for (k in 1:10) {
    pb$tick()
    
    train_fold <- train_data[folds != k, ]
    val_fold   <- train_data[folds == k, ]
    
    rf_model <- ranger(
      Sale_Price_log ~ ., 
      data = train_fold,
      num.trees = ntree,
      importance = "impurity",
      respect.unordered.factors = "order"
    )
    
    # log price prediction and transformation into raw again
    log_preds <- predict(rf_model, data = val_fold)$predictions
    preds <- exp(log_preds) - 1 
    
    true <- exp(val_fold$Sale_Price_log) - 1
    
    rmse <- sqrt(mean((preds - true)^2))
    mae  <- mean(abs(preds - true))
    mape <- mean(abs((true - preds) / true)) * 100
    
    cv_results <- rbind(cv_results, data.frame(
      num.trees = ntree,
      fold = k,
      RMSE = rmse,
      MAE = mae,
      MAPE = mape
    ))
  }
}

```

## Visualize Ranger's Random Forest results
```{r}
cv_long <- cv_results %>%
  pivot_longer(cols = c(RMSE, MAE, MAPE), names_to = "Metric", values_to = "Value")

# Plot all metrics
ggplot(cv_long, aes(x = factor(num.trees), y = Value)) +
  geom_boxplot(aes(fill = Metric), alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(
    title = "Manual 10-Fold CV: RMSE, MAE, and MAPE by Number of Trees",
    x = "Number of Trees",
    y = "Metric Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}
summary(comb_data$Sale_Price_Raw)
```

## XG Boosting
```{r}
library(xgboost)
library(dplyr)

# Reduce dataset to 20% for fast param search
set.seed(42)
train_subset <- train_data %>%
  sample_frac(0.2) %>%
  select(where(function(col) !(is.factor(col) && nlevels(col) > 100)))

# Create folds
n <- nrow(train_subset)
folds <- sample(rep(1:10, length.out = n))

# Define parameter grid
params_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1),
  max_depth = c(4, 6, 8),
  subsample = c(0.7, 0.8, 1.0),
  colsample_bytree = c(0.7, 0.8, 1.0),
  gamma = c(0, 1, 5),
  num.trees = c(100, 200, 300)
)

# Init results container
cv_results <- data.frame(
  eta = double(),
  max_depth = integer(),
  subsample = double(),
  colsample_bytree = double(),
  gamma = double(),
  num.trees = integer(),
  fold = integer(),
  RMSE = double(),
  MAE = double(),
  MAPE = double()
)

# Loop over param combinations
for (i in seq_len(nrow(params_grid))) {
  params <- params_grid[i, ]
  cat(sprintf("\nEvaluating combination %d of %d\n", i, nrow(params_grid)))
  
  for (k in 1:10) {
    cat("  Fold", k, "\n")
    
    train_fold <- train_subset[folds != k, ]
    val_fold   <- train_subset[folds == k, ]

    train_matrix <- model.matrix(Sale_Price_log ~ . - 1, data = train_fold)
    val_matrix   <- model.matrix(Sale_Price_log ~ . - 1, data = val_fold)

    y_train <- train_fold$Sale_Price_log
    y_val   <- val_fold$Sale_Price_log

    dtrain <- xgb.DMatrix(data = train_matrix, label = y_train)
    dval   <- xgb.DMatrix(data = val_matrix)

    xgb_model <- xgboost(
      data = dtrain,
      nrounds = params$num.trees,
      eta = params$eta,
      max_depth = params$max_depth,
      subsample = params$subsample,
      colsample_bytree = params$colsample_bytree,
      gamma = params$gamma,
      objective = "reg:squarederror",
      verbose = 0
    )

    log_preds <- predict(xgb_model, newdata = dval)
    preds <- exp(log_preds) - 1
    true  <- exp(y_val) - 1

    rmse <- sqrt(mean((preds - true)^2))
    mae  <- mean(abs(preds - true))
    mape <- mean(abs((true - preds) / true)) * 100

    cv_results <- rbind(cv_results, data.frame(
      eta = params$eta,
      max_depth = params$max_depth,
      subsample = params$subsample,
      colsample_bytree = params$colsample_bytree,
      gamma = params$gamma,
      num.trees = params$num.trees,
      fold = k,
      RMSE = rmse,
      MAE = mae,
      MAPE = mape
    ))

    rm(train_matrix, val_matrix, dtrain, dval, train_fold, val_fold)
    gc()
  }
}


```

## Visualize XG Boosting results
```{r}
cv_results_ <- cv_results %>%
  mutate(
    param_combo = paste0(
      "eta=", eta,
      ",depth=", max_depth,
      ",subsample=", subsample,
      ",colsample=", colsample_bytree,
      ",gamma=", gamma
    )
  )

library(GGally)

best_configs <- cv_results_ %>%
  group_by(eta, max_depth, subsample, colsample_bytree, gamma) %>%
  summarise(
    RMSE = mean(RMSE),
    MAE = mean(MAE),
    MAPE = mean(MAPE),
    .groups = "drop"
  )

GGally::ggparcoord(
  data = best_configs,
  columns = 6:8,  # RMSE, MAE, MAPE
  groupColumn = 1,  # e.g. eta
  scale = "globalminmax",
  showPoints = TRUE,
  title = "Parallel Coordinates of Metric Scores by Hyperparams"
) +
  theme_minimal()

```

```{r}
cv_results_ <- cv_results %>%
  mutate(
    param_combo = paste0(
      "eta=", eta,
      ",depth=", max_depth,
      ",subsample=", subsample,
      ",colsample=", colsample_bytree,
      ",gamma=", gamma
    )
  )

best_configs <- cv_results_ %>%
  group_by(param_combo, eta, max_depth, subsample, colsample_bytree, gamma, num.trees) %>%
  summarise(
    mean_RMSE = mean(RMSE),
    mean_MAE = mean(MAE),
    mean_MAPE = mean(MAPE),
    .groups = "drop"
  ) %>%
  arrange(mean_RMSE)  # You can switch to mean_MAE or mean_MAPE if you prefer

cat("Top 3 configurations based on RMSE:\n")
print(head(best_configs, 3))


```

```{r}
library(dplyr)
best_rmse_params <- cv_results %>%
  group_by(eta, num.trees) %>%
  summarise(mean_rmse = mean(RMSE), .groups = "drop") %>%
  arrange(mean_rmse) %>%
  dplyr::slice(1)

print(best_rmse_params)

library(dplyr)

# Best by MAE
best_mae <- cv_results %>%
  group_by(eta, num.trees) %>%
  summarise(mean_mae = mean(MAE), .groups = "drop") %>%
  arrange(mean_mae) %>%
  dplyr::slice(1)

print("Best configuration by MAE:")
print(best_mae)

# Best by MAPE
best_mape <- cv_results %>%
  group_by(eta, num.trees) %>%
  summarise(mean_mape = mean(MAPE), .groups = "drop") %>%
  arrange(mean_mape) %>%
  dplyr::slice(1)

print("Best configuration by MAPE:")
print(best_mape)

```

# Train XG Boosting & Ranger Random Forest
```{r}
library(xgboost)
library(Matrix)
library(dplyr)
library(progress)

# Remove target variables from features
train_features <- train_data %>% select(-Sale_Price_log, -Sale_Price_Raw)
test_features  <- test_data %>% select(-Sale_Price_log, -Sale_Price_Raw)

# Combine for consistent encoding
combined_matrix_data <- rbind(train_features, test_features)
sparse_matrix <- sparse.model.matrix(~ . -1, data = combined_matrix_data)

# Split into train and test matrices
train_matrix <- sparse_matrix[1:nrow(train_data), ]
test_matrix  <- sparse_matrix[(nrow(train_data) + 1):nrow(combined_matrix_data), ]

# Labels
train_label <- train_data$Sale_Price_log
test_label  <- test_data$Sale_Price_log

# Create DMatrix
dtrain <- xgb.DMatrix(data = train_matrix, label = train_label)
dtest  <- xgb.DMatrix(data = test_matrix, label = test_label)

# Predefine model params
best_params <- list(
  eta = 0.1,
  max_depth = 8,
  subsample = 1.0,
  colsample_bytree = 1.0,
  gamma = 0,
  alpha = 0.5,
  lambda = 1.0,
  objective = "reg:squarederror"
)
```

```{r}
# Run final training with early stopping and callback
xgb_model <- xgb.train(
  params = best_params,
  data = dtrain,
  nrounds = 300, 
  watchlist = list(train = dtrain),
  verbose = 1
)
#cat("Stopped at best_iteration =", cv$best_iteration, "\n")
```

# Test the final model
```{r}
# Predict
log_preds <- predict(xgb_model, dtest)
preds <- exp(log_preds) - 1
true  <- exp(test_label) - 1

# Metrics
rmse <- sqrt(mean((preds - true)^2))
mae  <- mean(abs(preds - true))
mape <- mean(abs((true - preds) / true)) * 100

cat(sprintf("Final XGBoost Performance on Test Set:\nRMSE: %.2f\nMAE: %.2f\nMAPE: %.2f%%\n",
            rmse, mae, mape))
```
# Plot the prediction vs actual
```{r}
plot(true, preds, pch = 16, col = rgb(0,0,0,0.3),
     main = "Predicted vs Actual", xlab = "True Price", ylab = "Predicted Price")
abline(0, 1, col = "red")
```

# Plot residuals vs predicted, distribution of rmse, and percentage error by segment
```{r}
residuals <- true - preds
df_resid <- data.frame(true, preds, residuals, error_pct = residuals / true * 100)

# 1a. Residual vs Prediction plot
library(ggplot2)
ggplot(df_resid, aes(x = preds, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Predicted", x = "Predicted", y = "Residual")

# 1b. Residual distribution
ggplot(df_resid, aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Residuals", x = "Residual", y = "Count")

# 1c. By segment – e.g., by number of bedrooms
ggplot(df_resid %>% mutate(bin = cut(preds, breaks = 5)),
       aes(x = bin, y = error_pct)) +
  geom_boxplot() +
  labs(title = "Percentage Error by Prediction Bin", x = "Predicted Price Bin", y = "% Error")

```
# RMSE by bedroom count
```{r}
# Assuming 'Bedrooms_Raw' and other features available in test_data
df_resid$Bedrooms_Raw <- test_data$Bedrooms_Raw

library(dplyr)
df_segment <- df_resid %>%
  group_by(Bedrooms_Raw) %>%
  summarise(mean_error = mean(residuals),
            rmse = sqrt(mean(residuals^2)),
            mae = mean(abs(residuals)),
            n = n()) %>%
  filter(n >= 50)

ggplot(df_segment, aes(x = factor(Bedrooms_Raw), y = rmse)) +
  geom_col(fill = "coral") +
  labs(title = "RMSE by Bedroom Count", x = "# Bedrooms", y = "RMSE")

```

# Bootstrap for Confidence Intervals
```{r}
nboot <- 50
pred_mat <- replicate(nboot, {
  idx <- sample(seq_len(nrow(train_matrix)), replace = TRUE)
  dboot <- xgb.DMatrix(data = train_matrix[idx, ], label = train_label[idx])
  model_boot <- xgboost(params = best_params, data = dboot, nrounds = 300, verbose = 1)
  predict(model_boot, dtest)
})

# For each test sample, get 2.5% and 97.5% quantile
ci_lower <- exp(apply(pred_mat, 1, quantile, probs = 0.025)) - 1
ci_upper <- exp(apply(pred_mat, 1, quantile, probs = 0.975)) - 1
```

```{r}
df_test <- data.frame(
  true = true,
  pred = preds,
  lower = ci_lower,
  upper = ci_upper
)

ggplot(df_test, aes(x = true, y = pred)) +
  geom_ribbon(aes(x = true, ymin = lower, ymax = upper), fill = "skyblue", alpha = 0.4) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(
    title = "Predicted vs True with 95% Bootstrap Confidence Intervals",
    x = "True Price",
    y = "Predicted Price (median)"
  ) +
  theme_minimal()
```

```{r}
mean_log_preds <- rowMeans(pred_mat)
mean_preds     <- exp(mean_log_preds) - 1
true_vals      <- exp(test_label) - 1

residuals <- mean_preds - true_vals
MAE   <- mean(abs(residuals))
RMSE  <- sqrt(mean(residuals^2))
MAPE  <- mean(abs(residuals / true_vals)) * 100

cat(sprintf("Bootstrap Test Set Metrics:\nMAE: %.2f\nRMSE: %.2f\nMAPE: %.2f%%\n", MAE, RMSE, MAPE))
```



